---
title: "Analysis of Mass Incarceration in the United States"
author: "Elizabeth Deng"
date: '2022-11-26'
output: html_document
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library("tidyverse")
library("knitr")
library("ggplot2")
```

```{r echo=FALSE}
source("../source/a4-helpers.R")
source("../source/analysis.R")
```

## Abstract 
This data analysis project looks at statistics from the Vera Institute 
regarding incarceration rates in America. The dataset spans several decades and 
all 50 states, and we will use it to depict trends in incarceration demographics  
over time. 


## Introduction 
>"The United States is home to less than 5 percent of the world’s population, 
> yet nearly 16 percent of all incarcerated people—roughly two million people—
> are held in our jails and prisons. We incarcerate more people per capita than 
> any country in the world."
> *Source: [Vera Institute](https://www.vera.org/ending-mass-incarceration)*  

This data set addresses the pressing social issue of mass incarceration and its  
effects on communities— specifically, and disproportionately, on impoverished 
communities and communities of color. Factors including (but not limited to):
over-policing of poor areas, the monetary bail system, racial disparities in 
stops and arrests, and a weakened public defender system *all* contribute to the 
matrix of domination being upheld. An unjust legal and incarceration system 
maintains a power structure, and this power structure is what the Vera Institute 
and its dataset aims to challenge.  

second paragraph about the questions I answer Doggo ipsum thicc borkf snoot, very
jealous pupper. Wrinkler bork maximum borkdrive wrinkler corgo shibe the 
neighborhood pupper, extremely cuuuuuute long bois ruff puggorino. Doge heck blop
heckin, long doggo. Doge you are doin me a concern big Doggo ipsum thicc borkf
snoot, very jealous pupper. Wrinkler bork maximum borkdrive wrinkler corgo shibe
the neighborhood pupper, extremely cuuuuuute long bois ruff puggorino. Doge
heck blop heckin, long doggo. Doge you are doin me a concern big ol fat boi 
doing me a 

## Data Summary 

```{r echo=FALSE}
source("../source/analysis.R")
```

I began my data analysis by first narrowing the dataset down to only Washington 
state. I wanted to look at the growth of the jail and prison populations in 
comparison to the growth of the overall population.

Between the years 1995 and 2016, Washington state had a total population growth 
of **`r wapop_change`%**. However, the population of individuals in jail grew by 
**`r jailpop_change`%** and the population of individuals in prison grew by 
**`r prisonpop_change`**%.

Then I broadened the analysis to the entirety of the United States. The most 
recent year that had the most available data was 2016, so the analysis focuses 
on that year.

Across all 50 states, the average white population was
**`r avg_whitepop`**%, the average black population was **`r avg_blackpop`**%,
and the average Latino population was **`r avg_latinxpop`**%. However, these
proportions are not the same in states' prison populations. The average 
percentage of prisoners who were white, black, and Latino were 
**`r avg_whiteprison`**%, **`r avg_blackprison`**%, and 
**`r avg_latinxprison`**% respectively.
We would expect the proportion of races to be about the same across the general
population and the prison population; however, fewer white people are imprisoned
at the same time that more black people are. 


## Growth of the U.S. Prison Population

```{r echo=FALSE}
plot_jail_pop_for_us()

```  

**CAPTION: The total population of individuals in U.S. jails over the course of
48 years.**  

***Analysis***  

The overarching questions regarding mass incarceration are: how has the state of incarceration changed over the years, and what could be the cause? The graph depicts a steep incline in jail population, beginning in around 1980. Between 1970 and 2018, the national jail population grew by over 500,000, which again prompts the question: why?  

## Growth of Prison Population by State  

```{r echo=FALSE}
plot_jail_pop_by_states(c("MA", "AZ", "TN", "CO", "AL", 
                          "MD", "WA"))  

```  

**CAPTION: A line graph depicting changes in jail population between 1970-2018, in Massachusetts, Arizona, Tennessee, Colorado, Alabama, Maryland, and Washington.**  

***Analysis***    

I chose these six states in particular because they have similar populations (around 7-8 million), which allows us to more clearly compare incarceration rate growth. One trend that stands out noticeably is a steady incline beginning in 1990, which peaks around 2008 and then begins to decline just a bit. This could represent quite a number of things, potentially including national justice reforms that reduced incarceration rates, or a sweeping impact caused by the 2008 recession. Another noticeable factor is that Tennessee's jail population begins and remains much higher than the other states', despite having similar population sizes.


